Parties, Models, Mobsters\newline {\fontsize{11}{15}\selectfont A New Implementation of Model-Based Recursive Partitioning in R}

### 2014-03-12 pkg/partykit/inst/NUS-2014/modelparty.Rnw
slides essentially ready except for class descriptions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@471 255f4249-2263-4417-b1ea-8498779e27a7

<<packages, echo=FALSE, results=hide>>=
library("partykit")
library("psychotools")
data("Topmodel2007", package = "psychotree")
options(digits = 4)
@

@@

\begin{itemize}
  \item Model-based recursive partitioning
  \begin{itemize}
    \item A generic approach
    \item Example: Bradley-Terry trees
  \end{itemize}
  \item Implementation in R
  \begin{itemize}
    \item Building blocks: Parties, models, mobsters
    \item Old implementation in \emph{party}
    \item All new implementation in \emph{partykit}
  \end{itemize}
  \item Illustration
\end{itemize}

@@

\textbf{Model-based algorithm}:

@@

@@

\textbf{Interpretation:} Captures instability for split-up into $C$ categories.

@@

@@

@@

\begin{frame}[fragile]
\frametitle{Illustration: Bradley-Terry trees}

@@

\begin{frame}[fragile]
\frametitle{Illustration: Bradley-Terry trees}

@@

\begin{frame}
\frametitle{Illustration: Bradley-Terry trees}

@@

\begin{itemize}
  \item All new implementation of model-based recursive partitioning in \emph{partykit}.
  \item Enables more efficient computations, rapid prototyping, flexible customization.
  \item Some new inference options.
\end{itemize}

@@

\small

@@

@@

Zeileis A, Hothorn T (2014).
 \textit{Parties, Models, Mobsters: A New Implementation of Model-Based Recursive Partitioning in R.}
 \code{vignette("mob", package = "partykit")}.

@@

>>>>

<<packages, echo=FALSE, results=hide>>=
library("partykit")
library("psychotools")
options(digits = 4)
@

<<data, echo=FALSE, results=hide>>=
data("Topmodel2007", package = "psychotree")
data("Titanic", package = "datasets")
ttnc <- as.data.frame(Titanic)
ttnc <- ttnc[rep(1:nrow(ttnc), ttnc$Freq), 1:4]
names(ttnc)[2] <- "Gender"
ttnc <- transform(ttnc, Treatment = factor(
  Gender == "Female" | Age == "Child",
  levels = c(FALSE, TRUE),
  labels = c("Normal\n(Male&Adult)", "Preferential\n(Female|Child)")
))
@

@@

\begin{itemize}
  \item Motivation: Trees and leaves
  \item Model-based (MOB) recursive partitioning
  \begin{itemize}
    \item Model estimation
    \item Tests for parameter instability
    \item Segmentation
    \item Pruning
    \item Local models
  \end{itemize}
  \item Implementation in R
  \begin{itemize}
    \item Building blocks: Parties, models, mobsters
    \item Old implementation in \emph{party}
    \item All new implementation in \emph{partykit}
  \end{itemize}
  \item Application
  \begin{itemize}
    \item Paired comparisons for Germany's Topmodel finalists
    \item Bradley-Terry trees
    \item Implementation from scratch
  \end{itemize}
\end{itemize}

@@

\textbf{Model-based (MOB) algorithm}:

@@

\begin{frame}[fragile]
\frametitle{Recursive partitioning}

\textbf{Example:} Logistic regression, assessing differences in
the effect of ``preferential treatment'' (``women and children first''?).

\textbf{Result:} Log-odds ratio of survival given treatment differs across
classes (slope), as does the survival probability of male adults (intercept).

\textbf{In R:} Generalized linear model tree with binomial family (and default logit link).

<<ttnc, results=hide>>=
mb <- glmtree(Survived ~ Treatment | Age + Gender + Class,
  data = ttnc, family = binomial, alpha = 0.05, prune = "BIC")
plot(mb)
print(mb)
@

\vspace*{-0.3cm}

\setkeys{Gin}{width=1.1\textwidth}
\hspace*{-0.4cm}%
<<ttnc-plot, fig=TRUE, height=7, width=12, echo=FALSE, results=hide>>=
plot(mb, tp_args = list(margins = c(1.5, 2, 1.5, 2.5)))
@

\begin{frame}[fragile]
\frametitle{Recursive partitioning}

<<ttnc-print, echo=FALSE>>=
ttnc$Treatment <- factor(ttnc$Treatment, labels = c("Normal", "Preferential"))
mb0 <- glmtree(Survived ~ Treatment | Age + Gender + Class,
  data = ttnc, family = binomial, alpha = 0.05, prune = "BIC")
mb0
@

@@

\textbf{Interpretation:} Capture instability for split-up into $C$ categories.

@@

\subsection{Local models}

\begin{frame}
\frametitle{Local models}

\textbf{Goals:}
\begin{itemize}
  \item Detection of interactions and nonlinearities in regressions.
  \item Add explanatory variables for models without regressors.
  \item Detect violations of parameter stability (measurement invariance)
    across several variables adaptively.
\end{itemize}

\textbf{Mobsters:}
\begin{itemize}
  \item Linear and generalized linear model trees (Zeileis \emph{et al.} 2008).
  \item Censored survival regression trees: parametric proportional hazard
    and accelerated failure time models (Zeileis \emph{et al.}~2008).
  \item Beta regression trees (Gr\"un \emph{et al.}~2012).
  \item Bradley-Terry trees for paired comparisons (Strobl \emph{et al.}~2011).
  \item Item response theory (IRT) trees: Rasch, rating scale and partial credit model
    (Strobl \emph{et al.}~2014, Abou El-Komboz \emph{et al.}~2014).
\end{itemize}

@@

\begin{frame}[fragile]
\frametitle{Implementation: Class}

\textbf{Class:} `\code{modelparty}'.

FIXME

@@

\textbf{Task:} Preference scaling of attractiveness.

\textbf{Model:} Paired comparison via Bradley-Terry.
\begin{itemize}
  \item Paired comparisons of attractiveness for
    \emph{Germany's Next Topmodel~2007} finalists:
    Barbara, Anni, Hana, Fiona, Mandy, Anja.
  \item Survey with 192~respondents at Universit{\"a}t T{\"u}bingen.
  \item Available covariates: Gender, age, familiarty with the TV~show.
  \item Familiarity assessed by yes/no questions:
    (1)~Do you recognize the women?/Do you know the show?
    (2)~Did you watch it regularly?
    (3)~Did you watch the final show?/Do you know who won?
\end{itemize}

\textbf{Mobster:} \code{bttree()} in \emph{psychotree} (Strobl \emph{et al.}~2011).

\textbf{Here:} Use \code{mob()} directly to build model from scratch.

\begin{frame}[fragile]
\frametitle{Bradley-Terry trees}

@@

\begin{frame}[fragile]
\frametitle{Bradley-Terry trees}

@@

@@

\begin{itemize}
  \item Synthesis of parametric data models and algorithmic tree models.
  \item Based on modern class of parameter instability tests.
  \item Aims to minimize clearly defined objective function by greedy forward search.
  \item Can be applied general class of parametric models.
  \item Alternative to traditional means of model specification,
        especially for variables with unknown association.
  \item All new implementation in \emph{partykit}.
  \item Enables more efficient computations, rapid prototyping, flexible customization.
\end{itemize}

@@

\footnotesize

\textbf{Software:}

@@

Zeileis A, Hothorn T (2014).
 \textit{Parties, Models, Mobsters: A New Implementation of Model-Based Recursive Partitioning in R.}
 \texttt{vignette("mob",~package~=~"partykit")}.

@@

\textbf{Inference:}

Zeileis A, Hornik K (2007).
 \dquote{Generalized {M}-Fluctuation Tests for Parameter Instability.}
 \textit{Statistica Neerlandica},
 \textbf{61}(4), 488--508.
 \doi{10.1111/j.1467-9574.2007.00371.x}

Merkle EC, Zeileis A (2013).
 \dquote{Tests of Measurement Invariance without Subgroups: A Generalization of Classical Methods.}
 \textit{Psychometrika}, \textbf{78}(1), 59--82.
 \doi{10.1007/S11336-012-9302-4}

Merkle EC, Fan J, Zeileis A (2014).
 \dquote{Testing for Measurement Invariance with Respect to an Ordinal Variable.}
 \textit{Psychometrika}. Forthcoming.

\footnotesize

\textbf{Trees:}

@@

Gr\"un B, Kosmidis I, Zeileis A (2012).
  \dquote{Extended Beta Regression in R: Shaken, Stirred, Mixed, and Partitioned.}
  \textit{Journal of Statistical Software}, \textbf{48}(11), 1--25.
  URL~\url{http://www.jstatsoft.org/v48/i11/}

%% \medskip
%% 
%% Rusch T, Lee I, Hornik K, Jank W, Zeileis A (2013).
%%   \dquote{Influencing Elections with Statistics: Targeting Voters with Logistic Regression Trees.}
%%   \textit{The Annals of Applied Statistics}, \textbf{7}(3), 1612--1639.
%%   \doi{10.1214/13-AOAS648}

Strobl C, Kopf J, Zeileis A (2013).
  \dquote{A New Method for Detecting Differential Item Functioning in the Rasch Model.}
  \textit{Psychometrika}. Forthcoming.
  \doi{10.1007/s11336-013-9388-3}

Abou El-Komboz B, Zeileis A, Strobl C (2014).
  \dquote{Detecting Differential Item and Step Functioning with Rating Scale and Partial Credit Trees.}
  \textit{Technical Report 152}, Department of Statistics, Ludwig-Maximilians-Universit\"at M\"unchen.
  URL~\url{http://epub.ub.uni-muenchen.de/17984/}
